No general trend for multiple dimensions of functional diversity in bird and mammal communities

### 2024-05-29 paper/GEB Submission/manuscript_GEB.Rmd
more changes

  Main Conclusions: General temporal trends indicate that on the aggregate one type of functional shift is not more prevalant than the other across many taxa, biomes, and realms. At the study-level, the majority of studies showed no species or functional trends, with the remainder of the studies falling into broad categories of functional shifts independent of species richness, functional redundancy loss, and increasing species richness.

@@

Ecological communities are experiencing unprecedented changes as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011] and fundamental changes in some of the most well-studied systems [e.g. coral bleaching, @sully2019].
At the local scale however, species diversity tells a different story.
Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017; @li2020].
While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.

`r colorize("Using general trends derived from limited data as a diagnostic for the state of biodiversity is a topic of ongoing debate.", "blue")` Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
Many of these criticisms reflect limitations of ecological data on the whole, `r colorize("leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities", "blue")` [@dornelas2023; @primack2018].

Functional diversity offers a potentially powerful addition to species-based approaches for detecting and describing community change by providing a mechanistic link between species' response to environmental change (*response traits*) and the processes they perform (*effect traits*) [@lavorel2002; @mcgill2006; @suding2008].
By describing the functional trait space, functional diversity metrics capture the disproportionate impact of losses or gains of functionally unique species.
Functional diversity metrics `r colorize("may therefore illuminate joint responses from functionally similar species or communities undetectible by looking at species identity alone.", "blue")`

Broad-scale trends in functional diversity are the aggregation of local-scale responses to disturbance and persistent pressures like environmental change or the introduction of invasive species.
On acute timescales, impacts on communities can range from reduction in abundance across the trait space to complete gains or losses in some aspect of the trait space [@boersma2016; @larson2021].
As communities recover or adapt to persistent pressures, changes in the functional space are determined by community assembly processes and limited by the species available in the regional species pool [@green2022].
Lost species may be replaced by functionally similar species when available, or novel species may colonize niches newly created by changing environmental conditions.
Critically, similar shifts in the functional space may be the result of both positive and negative conservation scenarios.
For example, expansion of the functional trait space may be evidence of recovery from disturbance or the introduction of a functionally unique invasive species.

`r colorize("The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow species richness trends", "blue")` [@dornelas2023].
While `r colorize("loss of functional diversity", "blue")` is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], `r colorize("functional diversity may be maintained even when species are lost from a community", "blue")` [@daz2001; @petchey2009].
Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020].
And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @larsen2018; @matuoka2020].
On paleoecological time scales, `r colorize("the functional space", "blue")` shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with significant impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].
For some time periods, functional diversity appears to be maintained for substantial portions of geological time [@hedberg2022].
`r colorize("Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds", "blue")` [@barnagaud2017; @jarzyna2016] `r colorize("and ray-finned fishes, sharks, and rays", "blue")` [@trindade-santos2020].

`r colorize("We have stronger expectations for changes in the prevalence of some individual traits. For example, animal body size is expected to decrease as a result of climate change, a phenomena that has been documented in multiple taxa empirically and experimentally", "blue")` [@caruso2014; @forster2012; @huss2019; @sheridan2011; @tseng2018].
`r colorize("There is some evidence that this holds true for birds ", "blue")`[@vanbuskirk2010]`r colorize(", but the picture is likely more complicated for mammals, where urbanization may actually lead to larger body sizes due to novel food sources", "blue")` [@hantak2021] `r colorize(", even as megafaunal loss leads to on average smaller community body size. For dietary traits, recent work documenting insect declines", "blue")` [@wagner2021] `r colorize("points to potentially significant negative impacts on insectivorous animals", "blue")` [@canaday1996; @sekercioglu2002; @tallamy2021].
`r colorize("Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species ", "blue")`[@cooke2019].`r colorize("Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems", "blue")`[@bain2020]`r colorize(", and loss of neotropical understory foragers even in protected areas", "blue")` [@pollock2022].

@@

Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at general, categorical, and study-level local trends in functional diversity in mammal and bird communities.
Our work complements efforts on a similar scale assessing temporal change in species and phyllogenetic diversity across taxa [@daru2021; @dornelas2014; @li2020; @vellend2013l; @vellend2017] or functional diversity change for a single taxa [@barnagaud2017; @jarzyna2016; @trindade-santos2020].
`r colorize("We did not detect an overall trend in any functional diversity metric, corrected or uncorrected. As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing dissimilarity through time", "blue")` [@dornelas2018]`r colorize(", indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends.", "blue")`

@@

`r colorize("Surprisingly, we found no evidence for some of the strongest predictions for anthropogenic impact. For example, we found no trend for mean body mass in bird or mammal communities. In fact, sensitivity results indicated that after excluding the shortest time series there is evidence of an increasing trend in body size. This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore restablishment are overshadowing climate change impacts, or current shifts are happening at an intraspecific level not captured by our data. Additionally, the majority of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
`r colorize("Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened", "blue")` [@ripple2015].

`r colorize("We similarly did not find evidence of changes in trait distributions predicted by previous global change work. We found no changes in the prevalence of different foraging strategies, despite documented losses of understory birds in the neotropics and some evidence that agricultural incursion particularly threatens arboreal species. Those shifts may therefore be the result of specific contexts and not generalizable to bird communities across the globe. Rather than declines in insectivory and increases in invertivory predicted by changes in resource availability, we instead found changes in the degree of consumption of fish and vertebrates. Neither increasing trends for fish consumption nor decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.", "blue")`

@@

`r colorize("While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact. In fact, study-level trends indicate quite the opposite, that functional shifts with negative or yet unknown implications for ecosystem processes may be going undetected by common species-based approaches. For example, loss of evenness in communities with constant species richness may be an first sign of a community being impacted by environmental change, with negative implications for stability and function.", "blue")`

>>>>

  Main Conclusions: General temporal trends indicate that on the aggregate one type of functional change is not more prevalant than the other across many taxa, biomes, and realms. At the study-level, the majority of studies showed no species or functional trends, with the remainder of the studies falling into broad categories of functional changes independent of species richness, functional redundancy loss, and increasing species richness.

@@

Ecological communities are experiencing unprecedented changes as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011] and fundamental changes in some of the most well-studied systems [e.g. coral bleaching, @sully2019].
At the local scale however, species diversity tells a different story.
Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017; @li2020].
While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change. `r colorize("Still, using general trends derived from limited data as a diagnostic for the state of biodiversity is a topic of ongoing debate.", "blue")` Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
Many of these criticisms reflect limitations of ecological data on the whole, `r colorize("leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities", "blue")` [@dornelas2023; @primack2018].

Functional diversity offers a potentially powerful addition to species-based approaches for detecting and describing community change by capturing changes in the traits that underpin fundamental ecological processes [@lavorel2002; @mcgill2006; @suding2008].
Functional diversity metrics capture complimentary aspects of the trait space including the volume of the multidimensional space (functional diversity), how species and abundance are distributed across the space (functional evenness and divergence), and where the space occurs across the range of possible trait distributions (community weighted trait means).
Considered together, they describe the impacts of changes in species identity on the building blocks of stability, function, and resilience.

Broad-scale trends in functional diversity are the aggregation of local-scale responses to acute pulse disturbance, persistent press disturbances, and periods of relative stasis [@inamine2022].
Pulse disturbances have a sudden one time impact on the community, changing abundance distributions (functional evenness or divergence) or reducing the volume of the functional space [@boersma2016; @hillebrand2020; @larson2021]. While press disturbances can also lead to loss of functional volume as species cannot adapt rapidly to new conditions, it can also lead to wholesale shifts in the location of the functional space as species fill in new environmental space.
Rather than identifying outcomes of one of these specific processes, observational time series of functional diversity therefore capture the cumulative affects of multiple potentially competing processes.

`r colorize("The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow species richness trends", "blue")` [@dornelas2023; @larson2021].
`r colorize("While loss of functional diversityis frequently cited as one of the most pressing concerns of the anthropocene ", "blue")`[@cardinale2012; @dirzo2014; @young2016], observational trends show significant functional loss in some studies [@flynn2009] and no loss even in the most heavily impacted communities for others [@edwards2013; @larsen2018; @matuoka2020]. `r colorize("Broad-scale studies are limited, but show for example functional richness increases for North American birds", "blue")` [@barnagaud2017; @jarzyna2016]. `r colorize("We have stronger expectations for changes in the prevalence of some individual traits. For example, animal body size is expected to decrease as a result of climate change, a phenomena that has been documented in multiple taxa empirically and experimentally", "blue")` [@caruso2014; @forster2012; @huss2019; @sheridan2011; @tseng2018]. `r colorize("For dietary traits, recent work documenting insect declines", "blue")` [@wagner2021] `r colorize("points to potentially significant negative impacts on insectivorous animals", "blue")` [@canaday1996; @sekercioglu2002; @tallamy2021].
`r colorize("Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species ", "blue")`[@cooke2019].`r colorize("Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems", "blue")`[@bain2020]`r colorize(", and loss of neotropical understory foragers even in protected areas", "blue")` [@pollock2022].

@@

Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at general, categorical, and study-level local trends in functional diversity in mammal and bird communities.
Our work complements efforts on a similar scale assessing temporal change in species and phyllogenetic diversity across taxa [@daru2021; @dornelas2014; @li2020; @vellend2013; @vellend2017] or functional diversity change for a single taxa [@barnagaud2017; @jarzyna2016; @trindade-santos2020].
`r colorize("We did not detect an overall trend in any functional diversity metric, corrected or uncorrected. As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing dissimilarity through time", "blue")` [@dornelas2018]`r colorize(", indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends.", "blue")`

@@

`r colorize("Surprisingly, we found no evidence for some of the strongest predictions for anthropogenic impact. For example, we found no trend for mean body mass in bird or mammal communities. In fact, sensitivity results indicated that after excluding the shortest time series there is evidence of an increasing trend in body size. This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore restablishment are overshadowing climate change impacts", "blue") `[@hantak2021]`r colorize(", or current shifts are happening at an intraspecific level not captured by our data. Additionally, the majority of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
`r colorize("Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened", "blue")` [@ripple2015].

`r colorize("We similarly did not find evidence of changes in trait distributions predicted by previous global change work. We found no changes in the prevalence of different foraging strategies, despite documented losses of understory birds in the neotropics and some evidence that agricultural incursion particularly threatens arboreal species. Those shifts may therefore be the result of specific contexts and not generalizable to bird communities across the globe. Rather than declines in insectivory and increases in invertivory predicted by changes in resource availability, we instead found changes in the degree of consumption of fish and vertebrates. Neither increasing trends for fish consumption nor decreasing trends for vertebrate consumption in marine and tropical studies are obviously consistent with known changes in those contexts and warrant further examination.", "blue")`

@@

`r colorize("While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact. In fact, study-level trends indicate quite the opposite, that functional changes with negative or yet unknown implications for ecosystem processes may be going undetected by common species-based approaches. For example, loss of evenness in communities with constant species richness may be an first sign of a community being impacted by environmental change, with negative implications for stability and function.", "blue")`
